---
title: "Cyclistic Bike-Share Analysis"
output:
  pdf_document: default
  word_document: default
  html_document:
    date: "2021/12/15"
pagetitle: Cyclistic Bike-Share Analysis
---
```{r Installation}
local({r <- getOption("repos")
       r["CRAN"] <- "http://cran.r-project.org" 
       options(repos=r)
})
install.packages("tidyverse")
library(tidyverse)
install.packages("janitor")
install.packages("lubridate")
library("janitor")
library(lubridate)
install.packages("skimr")
library(skimr)
library(dplyr)
install.packages("ggplot2")
library(ggplot2)
```

#=====================
# STEP 1: COLLECT DATA
#=====================

```{r Upload Divvy datasets (csv files) here}
data_202012 <-read.csv("202012-divvy-tripdata.csv")
data_202101 <-read.csv("202101-divvy-tripdata.csv")
data_202102 <-read.csv("202102-divvy-tripdata.csv")
data_202103 <-read.csv("202103-divvy-tripdata.csv")
data_202104 <-read.csv("202104-divvy-tripdata.csv")
data_202105 <-read.csv("202105-divvy-tripdata.csv")
data_202106 <-read.csv("202106-divvy-tripdata.csv")
data_202107 <-read.csv("202107-divvy-tripdata.csv")
data_202108 <-read.csv("202108-divvy-tripdata.csv")
data_202109 <-read.csv("202109-divvy-tripdata.csv")
data_202110 <-read.csv("202110-divvy-tripdata.csv")
data_202111 <-read.csv("202111-divvy-tripdata.csv")
```

#====================================================
# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
#====================================================

```{r Compare column names each of the files, echo=TRUE, message=FALSE, warning=FALSE}
colnames(data_202012)
colnames(data_202101)
colnames(data_202102)
colnames(data_202103)
colnames(data_202104)
colnames(data_202105)
colnames(data_202106)
colnames(data_202107)
colnames(data_202108)
colnames(data_202109)
colnames(data_202110)
colnames(data_202111)
```

```{r Inspect the dataframes and look for incongruencies}
str(data_202012)
str(data_202101)
str(data_202102)
str(data_202103)
str(data_202104)
str(data_202105)
str(data_202106)
str(data_202107)
str(data_202108)
str(data_202109)
str(data_202110)
str(data_202111)
```

```{r Convert ride_id and rideable_type to character so that they can stack correctly}
data_202012 <-  mutate(data_202012, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
data_202101 <-  mutate(data_202101, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
data_202102 <-  mutate(data_202102, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
data_202103 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202104 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202105 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202106 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202107 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202108 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202109 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202110 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
data_202111 <-  mutate(data_202102, ride_id = as.character(ride_id)
                       ,rideable_type = as.character(rideable_type)) 
```

```{r Stack individual quarter data frames into one big data frame}
all_trips <- bind_rows(data_202012
                       , data_202101
                       , data_202102
                       , data_202103
                       , data_202104
                       , data_202105
                       , data_202106
                       , data_202107
                       , data_202108
                       , data_202109
                       , data_202110
                       , data_202111)
```

```{r Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```            

```{r Rename columns}
all_trips <- all_trips %>% rename(trip_id= ride_id
                                  ,ride_type= rideable_type
                                  ,start_time= started_at
                                  ,end_time= ended_at
                                  ,from_station_name = start_station_name 
                                  ,from_station_id = start_station_id
                                  ,to_station_name = end_station_name
                                  ,to_station_id =  end_station_id)
       
```

#======================================================
# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
#======================================================
```{r  Inspect the new table that has been created}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips) 
```

```{r  Data Cleaning}
table(all_trips$member_casual)
```

```{r  Reassign to the desired values (we will go with the current 2020 labels)}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

```{r  Check to make sure the proper number of observations were reassigned)}
table(all_trips$member_casual)
```

```{r  Add columns that list the date, month, day, and year of each ride}
all_trips$date <- as.Date(all_trips$start_time)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

```{r  Add a "ride_length" calculation to all_trips (in seconds)}
all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)
```

```{r  Inspect the structure of the columns}
str(all_trips)
```

```{r  Convert ride_length from Factor to numeric so we can run calculations on the data}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

```{r  Remove bad data}
all_trips_v2 <- all_trips[!(all_trips$from_station_name  == "HQ QR" | all_trips$ride_length<0),]
```

#=====================================
# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
#=====================================
```{r  Descriptive analysis on ride_length (all figures in seconds)}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
summary(all_trips_v2)
```

```{r   You can condense the four lines above to one line using summary() on the specific attribute}
summary(all_trips_v2$ride_length)
```

```{r   Compare members and casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

```{r   See the average ride time by each day for members vs casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r   Notice that the days of the week are out of order. Need fix that.}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("星期日", "星期一", "星期二", "星期三", "星期四", "星期五", "星期六"))
```

```{r   Run the average ride time by each day for members vs casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r   Visualize the number of rides by rider type}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

```{r   Create a visualization for average duration}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

#=================================================
# STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
#=================================================
```{r    Create a csv file}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, "avg_ride_length.csv")
write.csv(all_trips_v2, "tripdata.csv")